How to publish in Biological Sciences: a guide for the uninitiated

### 2021-07-31 1.07-citations.Rmd
Add files via upload
# after 1st complete read and edit

Citations matter as they reflect the number of people who found your research useful. Getting your research cited is not always straightforward. We have already mentioned here about publishing your research in journals with higher Impact Factors. This will likely help you get your research read. But there are lots of other tactics and here I have produced a list although I imagine it is not exhaustive.

## What are your citation metrics?
A number of different databases compile metrics which you can access. In addition, they compile a bunch of extra performance metrics that you would have to pay to access, and in general these are available to a select number of recruitment staff at your institution. Here I will talk about Google Scholar, Scopus and the Web of Science (see [an article on their differences here](http://howtowriteaphd.com/databases.html)). In summary, for Biological Sciences, the Web of Science and scopus are likely to give you very similar statistics. Scopus favours social-sciences a little more, so if your subject area does cross-over then you might see some extra cites there. Google Scholar is going to give you the highest metrics, but it doesn't mean that they are particularly false (although this is certainly the not the best curated database). The joy of Google Scholar is that it is much more cosmopolitan on the journals and theses that it includes. So it actually gives you a better idea of who is using your research. It will also pick up grey literature and predatory publishers, so you should interpret the outputs with care.

Each of these databases offers you the opportunity of curating your own author profile that (in the case of Scopus and WoS) will link with your ORCID, and (in the case of WoS) link with your Publons account. Your institution may want you to actively curate your profile to help them with assessing your institution's metrics. It is surprising how bad these are when left to the bots, especially if you have your name recorded in more than one way.I also think that it's worthwhile creating and curating a Google Scholar profile. As an editor, this is one of the easiest ways of looking up an author or potential reviewer. For anyone that uses Google Scholar, your name will then be underlined, and they can quickly find your other published works. Google Scholar also indexes preprints, which may be a boon to you as an Early Career Researcher.

Because all of these indices are cumulative, you can expect that anyone who started their career before you will have higher numbers. It is also worth being aware that some fields get higher citations because there are more people working in them, and therefore more articles being published.

### Total Citations
Your total number of citations is simply the total number of times that anyone has cited one of your works within the respective database. Note that, the cited article itself doesn't have to appear in the database in order to be counted in total cites, but (clearly) the article that cites it must. For this reason, you may well have a different number.

@@

#### Normalisaing the H-index and other citation metrics
The H-index (and other indices) are sometimes normalised for field of study (fixed - dynamic; broad - narrow), types of citing documents and year of publication [see @ioannidis2016citation]. You can expect to see some of these field normalisations, for example by [Dimensions](https://badge.dimensions.ai/) (see Figure \@ref(fig:Dimensions) who provide you with, Total citations (as the name implies), Recent citations (citations within the last 2 years), Field citation ratio (comparing the article to ones of a similar age in in a similar field), and the Relative citation ratio (relative to 1.0, this gives you an idea of how well cited the article is compared to others in its area of research). As an Early Career Researcher, you may feel that your H-index is not particularly impressive (Table \@ref(tab:cite)), in which case, you may want to consider some of these normalised metrics if they show your articles in a much more favourable light.

@@

### G-index
The G-index is similar to the H-index, but places more emphasis on the actual number of citations in your top cited articles [@egghe2006theory]. To calculate this, you will need to organise your publications by the number of times that they have been cited. For ease of explanation, start with your H-index, and add up citations to all of the preceding articles (1st, 2nd, 3rd, 4th & 5th). In order for your G-index to be 5, all citations up to 5 will need to be more than or equal to 25 (5^2^). For your G-index to increase to 6, the sum of all citations to your first 6 papers will need to be equal or greater than 36 (6^2^).

Because the G-index uses all citations in your top cited articles, not simply the minimum used to obtain the H-index, your G-index should always (at least) be equal to your H-index and will often be higher. This is because of the phenomenon that articles that get cited more, continue by getting increasing numbers of citations (see [below](#winnertakesall)) Therefore, you may favour using the G-index if your top papers are particularly well cited.

### R-index
Your R-index is a measure of your active network [@ioannidis2008measuring]. It is rather like the H-index it that it records the count of the number of people that you've collaborated with on a number of papers. But these need to be the same people. This is then divided by your total number of papers. For example, if your thesis produced 5 papers with the same 5 authors and this is your total number of papers, your R-index will be 1. However, if two of the papers lacked one of the people, your R-index will drop to 1.66. If you don't work with those 5 other people again, then your R-index will grow continuously until you have another productive network. See [Part I](#networks) for a fuller explanation of the R-index and the importance of networks.

### a-index
The a-index is calculated from the total number of citations divided by H^2^ [@sidiropoulos2007generalized]. In the example given in Table \@ref(tab:cite), Total citations is: 97. H^2^ is 25 (5^2^), and so the a-index is 97/25 = 3.88. This index gives a measure of how top-heavy someone's citations are. The a-index gets larger as the highest ranked papers receive a disproportionate amount of citations. This is known as the [Matthew Effect](#winnertakesall), and this is a general feature of a lot of researchers' citation ranks.

### i10-index
Your i10 index represents the number of articles that you have written which have been cited 10 times or more in that database. The i10 is often expressed over a time frame (for example the last 5 years) which may well exclude papers that you write that have an immediate but not a lasting impact.

@@

There are also Altmetrics handed out as a means of showing the alternative impact for articles that you've written. These are explained in [Part III](altmetrics.html).

## How to increase your citations 
Increasing your total number of citations is likely to impact positively on all of your metrics, but this is not necessarily true, and hence why you may be assessed on more than one metric. In the following subsections, I go through some of the well known ways in which you can increase the readership of your paper, and with this the likelihood that someone will cite it.

### Media release
Some journals will want you to write a media release that they may use to promote your paper. You can also proactively [write a press release](pressrelease.html) and send it to your Media office at your institution. Writing a release for the media is very different from most of the writing that we've talked about so far and you can find a guide in a [chapter below](pressrelease.html).

### Social media
Social media is another way of getting your article to the attention of more people. Twitter has been adopted by many scientists and tweeting out an article that gets retweeted by the right people (i.e. social media hubs - with lots of followers) can get you tens or even hundreds of thousands of views. This is likely to be far in excess of anything your article will get by passively sitting on the journal website.

@@

The lesson followed swiftly. Within days of the online publication, the Tweets started rolling in, with a great many people very upset with what they saw as an insult to younger researchers who had a good social media following on Twitter [@woolston2014clash]. The social media eruption over this article made it score an [Altmetric](altmetrics.html) of 2489, ranked 1st of articles published at that time in that journal, and 9th in all journals. Although Hall had written the article in a light hearted way (poking fun at metrics and social media), the backlash was massive (although some did appreciate the joke). There are several lessons here: First, is that what seems like a joke to you may be offensive to others, especially those from other cultures. Second, is that items in social media can go badly wrong, and be taken in ways that you had never imagined. Third, is that social media goes well beyond academia - as is the intention with the dissemination, and that this may also have repercussions.

@@

#### Academic social media sites
Sites such as ResearchGate and Academia.edu will play a role in increasing citations to your work, especially if you have a large following. Rather like sharing on traditional social-media sites, the more people are exposed to your work, the more they are likely to read and cite it. In the case of academic social media sites, the audience is more niche and so more likely to find your work of interest.

### Popular articles
[Writing popular articles](populararticle.html) is a good way to ensure that your paper gets some media attention. There are outlets like [The Conversation](https://theconversation.com/) that specialise in publishing popular articles written by academics. But you might also find that academic societies that you belong to have newsletters that you can contribute a popular article to about your paper. As long as the circulation of these newsletters is big it should enhance the numbers of people that eventually read about your work.

@@

There has even been a call for another metric, the **s-index** which is the equivalent of the h-index but for self citations [@flatt2017improving]: "the total number of s papers that he or she has published that have at least the same amount of s self-citations".

How do you make sure that their citations are not gratuitous? Simply by only citing a work when you need to. Note that many citation indices will provide your citation scores both with and without self citations. As a rule of thumb, metrics suggest reasonable levels self-citations are ~30% [@aksnes2003macro]. In short, there is nothing wrong with citing yourself, but you should only do so when you need to.

### Citing papers submitted and in press
If you, or others, have papers that are 'in press', then you can cite these in a manuscript submitted to a journal for peer review. However, 'in press' does mean that they have been formally accepted by an editor at a journal. These days the concept of 'in press' is not quite the same as it was. It did used to take months before an accepted paper would appear in a journal issue. These days it is likely that you'll have an online version very quickly after acceptance. Hence, by the time you submit a manuscript with citations to papers that are 'in press', by the next round of revisions or acceptance, they should be fully available. If papers are not available by the time you set your proofs, then you should be asked to remove these citatoins. My impression is that many authors claim that a paper is 'in press' when the reality is that it's simply been submitted and that they are hoping it will be 'in press'.

You should not add citations to manuscripts that have simply been submitted to other journals. If you really need to, then submitting and then citing as a [preprint](preprint.html) is really your only option. Some editors will not accept citations of preprints. An exception to this might be when papers are submitted together to the same journal at the same time. If this is the case, then you can point this out to the editor in your [cover letter](letter.html).

If the contents of this manuscript are pivotal for understanding the submitted version, then you should make sure that there is a copy of the 'in press' or 'submitted' manuscript available for the reviewers. The easiest way of doing this will be as a [preprint](preprint.html). If this is not possible, then you may need to upload the ms as an additional file - but you should seek guidence from the journal editor.

@@

Special issues are often edited by people who have an interest in pulling together this kind of research. They may encourage other authors in the special issue to cross cite articles, and may write an editorial that includes citations to yours and other papers. Of course the best position to be in is when you are one of the organising members. Your name will be associated with the special issue as well as your contribution. This means that when the special issue is published, each of the papers in it will already have citations leading to a positive [Matthew Effect](#matthew) from day one of publication.

Probably the easiest way of pulling together a special issue is to organise a symposium at a conference and ask those participating to contribute to the special issue. If this is done long enough in advance and you are good at organising then it can work very well. Talk to your adviser about the possibility of their help in organising a symposium.

### Traditional Media 
To get your work cited you need to get people reading it. As scientists read the [traditional media](pressrelease.html) it can be a way of bringing your work to their attention, and result in additional citations. There are other benefits of having your work highlighted by the traditional media then you should always make sure that you get a plug in for your institution and any prominent funders. There are now additional metrics (e.g. [Altmetrics](altmetrics.html)) that track traditional and social media and in the future you might be able to use these to your advantage in getting a job, a promotion or tenure.

### Not so legitimate ways of increasing citations
As well as the legitimate ways of increasing citations and visibility of your work that I've listed above there are other illegitimate ways that I have heard about. I have heard of certain laboratories that have quid-pro-quo arrangements where they agree to cite each other's publications [see @ritchie2020science].

Another way of increasing citations might be to have a co-author that is very well known. Certainly having a co-author that is very well known is likely to increase your citations but only if they actually do something to contribute to the work. See comments about [ghost authorship above](authors.html).

@@

Journals have also been known to manipulate citations in order to increase their Impact Factors and therefore their perceived level of standing. A group of physics journals from Romania were found to have clearly manipulated self-citations to increase their own impact factors [see @heneberg2016excessive]. There can also be a level of coercive pressure from journal editors to cite papers from their journals published within the 2 year citation window [@chorus2016large-scale]. Such practices are now systematically analysed by the larger literature databases, and clear levels of [Impact Factor](impactfactor.html) manipulation result in delisting of the journal from the database ([see part 2](databases.html)).

@@

As you read through papers in your specialist area you will notice but there are some papers that seem to get cited over and over again as standard examples of a particular phenomenon. These are often some of the first examples published and also being published in high-ranking journals. If you are the 20th person to have shown a particular phenomenon then it is unlikely that your paper is going to end up being cited the most.

There have been investigations into what makes certain papers more likely to be cited further. One of the factors that appears to be significant is the "small-study effect" - low precision studies that produce large effect sizes, either spuriously (i.e. a Type I error: @forstmeier2017detecting, or genuinely @fanelli2017meta-assessment).

>>>>

Citations matter as they reflect the number of times people who found your research useful. Although you can look at the citations of each paper, your publications will most likely be summarised into a citation metric to get an overview of how your work is used. Here I provide a guide to some of these [metrics](#metrics), how they are calculated and used. Growing your citations by getting your research cited is not always straightforward. Publishing your research in journals with higher [Impact Factors](#impactfactor) will likely help you get your research read, but there are lots of other tactics and I give some suggestions for [increasing your citations](#increasingcitations).

## What are your citation metrics? {#metrics}
A number of different databases compile metrics which you can access. In addition, they compile a bunch of extra performance metrics that you would have to pay to access, and in general these are available to a select number of recruitment staff at your institution. Here I will talk about Google Scholar, Scopus and the Web of Science [see @martin-martin2018google; @martin-martin2021google on their differences]. In summary, for Biological Sciences, the Web of Science (WoS) and scopus are likely to give you very similar statistics. Scopus favours social-sciences a little more, so if your subject area does cross-over then you might see some extra citations there. Google Scholar is going to give you the highest metrics, but it doesn't mean that they are particularly false (although this is certainly the not the best curated database). The joy of Google Scholar is that it is much more cosmopolitan on the journals and theses that it includes. So it actually gives you a better idea of who is using your research. It will also pick up grey literature and predatory publishers, so you should interpret the outputs with care.

Each of these databases offers you the opportunity of curating your own author profile that (in the case of Scopus and WoS) will link with your ORCID, and (in the case of WoS) link with your Publons account. Your institution may want you to actively curate your profile to help them with assessing your institution's metrics. It is surprising how bad these are when left to the bots, especially if you have your name recorded in more than one way. I also think that it's worthwhile creating and curating a Google Scholar profile, and link it directly to your [website](#website). As an editor, this is one of the easiest ways of looking up an author or potential reviewer. For anyone that uses Google Scholar, your name will then be underlined, and they can quickly find your other published works. Google Scholar also indexes preprints, which may be a boon to you as an Early Career Researcher.

Because all of these indices are cumulative, you can expect that anyone who started their career before you will have higher numbers. It is also worth being aware that some fields get higher citations because there are more people working in them, and therefore more articles are being published.

### Total Citations
Your total number of citations is simply the total number of times that anyone has cited one of your works within the respective database. Note that, the cited article itself doesn't have to appear in the database in order to be counted in total cites, but (clearly) the article that cites it must. For this reason, you are likely to have a different number of total citations in different databases.

@@

#### Normalisaing the H-index and other citation metrics
The H-index (and other indices) are sometimes normalised for field of study (fixed - dynamic; broad - narrow), types of citing documents and year of publication [see @ioannidis2016citation]. You can expect to see some of these field normalisations, for example by [Dimensions](https://badge.dimensions.ai/) (see Figure \@ref(fig:Dimensions)) who provide you with, total citations (as the name implies), Recent citations (citations within the last 2 years), field citation ratio (comparing the article to ones of a similar age in in a similar field), and the Relative citation ratio (relative to 1.0, this gives you an idea of how well cited the article is compared to others in its area of research). As an Early Career Researcher, you may feel that your H-index is not particularly impressive (Table \@ref(tab:cite)), in which case, you may want to consider some of these normalised metrics if they show your articles in a more favourable light.

@@

### G-index
The G-index is similar to the H-index, but places more emphasis on the actual number of citations in your top cited articles [@egghe2006theory]. To calculate this, you will need to organise your publications by the number of times that they have been cited. For ease of explanation, start with your H-index, and add up citations to all of the preceding articles (1st, 2nd, 3rd, 4th & 5th). For example, in order for your G-index to be 5, all citations up to 5 will need to be more than or equal to 25 (5^2^). For your G-index to increase to 6, the sum of all citations to your first 6 papers will need to be equal or greater than 36 (6^2^). Our fictional researcher in Table \@ref(tab:cite) has a G-index of 9 as citations to the first 9 articles add up to 98 which is greater than 81 (i.e. >9^2^).

Because the G-index uses all citations in your top cited articles, not simply the minimum used to obtain the H-index, your G-index should always be equal to your H-index and will often be higher. This is because of the phenomenon that articles that get cited more, continue by getting increasing numbers of citations (known as the [Matthew Effect](#winnertakesall), see [below](#winnertakesall)). Therefore, you may favour using the G-index if your top papers are particularly well cited.

### R-index
Your R-index is a measure of your active network [@ioannidis2008measuring]. It is rather like the H-index it that it records the count of the number of people that you've collaborated with on a number of papers. But these need to be the same people. This is then divided by your total number of papers. For example, if your thesis produced 5 papers with the same 5 authors and this is your total number of papers, your R-index will be 1. However, if two of the papers lacked one of the people, your R-index will drop to 1.66. If you don't work with those 5 other people again, then your R-index will grow continuously until you have another productive network. See the section on [networks](#networks) for a fuller explanation of the R-index and the importance of networks.

### a-index
The a-index is calculated from the total number of citations divided by H^2^ [@sidiropoulos2007generalized]. In the example given in Table \@ref(tab:cite), Total citations is: 98. H^2^ is 25 (5^2^), and so the a-index is 98/25 = 3.92. This index gives a measure of how top-heavy someone's citations are. The a-index gets larger as the highest ranked papers receive a disproportionate amount of citations. This is known as the [Matthew Effect](#winnertakesall) (see [below](#winnertakesall)), and this is a general feature of a lot of researchers' citation ranks.

### i10-index
Your i10 index represents the number of articles that you have written which have been cited 10 times or more in that database. The i10 is often expressed over a time frame (for example the last 5 years) which may well exclude papers that you write that have an immediate but not a lasting impact. Moreover, looking at an index calculated over the last 5 years will let you know how active older colleagues are. Given that most indices are cumulative, older researchers almost always look better. But a time constrained index may provide additional insight into their relative productivity.

@@

There are also Altmetrics handed out as a means of showing the alternative impact for articles that you've written. These are explained in [Part III](#altmetrics).

## How to increase your citations {#increasingcitations}
Increasing your total number of citations is likely to impact positively on all of your metrics, but this is not necessarily true, and hence why you may be assessed on more than one metric. In the following subsections, I go through some of the well known ways in which you can increase the readership of your paper, and with this the likelihood that someone will cite it.

### Media release
Some journals will want you to write a media release that they may use to promote your paper. You can also proactively [write a press release](#pressrelease) and send it to the media office at your institution. Writing a release for the media is very different from scientific writing, and you can find a guide in [Part III](#pressrelease).

### Social media {#social}
Social media is another way of getting your article to the attention of more people. Twitter has been adopted by many scientists and tweeting out an article that gets retweeted by the right people (i.e. social media hubs - with lots of followers) can get you tens or even hundreds of thousands of views. This is likely to be far in excess of anything your article will get by passively sitting on the journal website.

@@

A social media backlash followed swiftly. Within days of the online publication, the Tweets started rolling in, with a great many people very upset with what they saw as an insult to younger researchers who had a good social media following on Twitter [@woolston2014clash]. The social media eruption over this article made it score an [Altmetric](#altmetrics) of 2489, ranked 1st of articles published at that time in that journal, and 9th in all journals. Although Hall had written the article in a light hearted way (poking fun at metrics and social media), the backlash was massive (although some did appreciate the joke).

There are several lessons here: First, is that what seems like a joke to you (including those immediately around you) may be offensive to others, especially those from other cultures. Second, is that items in social media can go badly wrong, and be taken in ways that you had never imagined. Third, is that social media goes well beyond academia - as is the intention with the dissemination, and that this may also have repercussions.

@@

#### Academic social media sites
Sites such as ResearchGate and Academia.edu will play a role in increasing citations to your work, especially if you have a large following. Rather like sharing on traditional social-media sites, the more people who are exposed to your work, the more likely they are to read and cite it. In the case of academic social media sites, the audience is more niche and so more likely to find your work of interest.

### Popular articles
[Writing popular articles](#populararticle) is a good way to ensure that your paper gets some media attention. There are outlets like [The Conversation](https://theconversation.com/) that specialise in publishing popular articles written by academics. But you might also find that academic societies that you belong to have newsletters that you can contribute a popular article to about your paper. As long as the circulation of these newsletters is big it should enhance the numbers of people that eventually read about your work.

@@

There has even been a call for another metric, the **s-index** which is the equivalent of the h-index but for self citations [@flatt2017improving]: "the total number of s papers that an author has published that have at least the same amount of s self-citations".

How do you make sure that their citations are not gratuitous? Simply by only citing a work when you need to. Note that many citation indices will provide your citation scores both with and without self citations. As a rule of thumb, metrics suggest reasonable levels self-citations in the biological sciences are ~30% [@aksnes2003macro]. In short, there is nothing wrong with citing yourself, but you should only do so when you need to.

### Citing papers submitted and in press
If you, or others, have papers that are 'in press', then you can cite these in a manuscript submitted to a journal for peer review. However, 'in press' does mean that they have been formally accepted by an editor at a journal. These days the concept of 'in press' is not quite the same as it was. It used to take months before an accepted paper would appear in a journal issue. These days it is likely that you'll have an online version very quickly after acceptance. Hence, by the time you submit a manuscript with citations to papers that are 'in press', by the next round of revisions or acceptance, they should be fully available. If papers are not available by the time you send back your proofs, then you may well be asked to remove these citations. My impression is that many authors claim that a paper is 'in press' when the reality is that it's simply been submitted and that they are hoping it will be 'in press'.

You should not add citations to manuscripts that have simply been submitted to other journals. If you really need to, then submit and cite as a [preprint](#preprint) is really your only option. Some editors will not accept citations of preprints. An exception to this might be when papers are submitted together to the same journal at the same time. If this is the case, then you can point this out to the editor in your [cover letter](#letter).

If the contents of this manuscript are pivotal for understanding the submitted version, then you should make sure that there is a copy of the 'in press' or 'submitted' manuscript available for the reviewers. The easiest way of doing this will be as a [preprint](#preprint). If this is not possible, then you may need to upload the ms as an additional file - but you should seek guidence from the journal editor.

@@

Special issues are often edited by people who have an interest in pulling together this kind of research. They may encourage other authors in the special issue to cross cite articles, and may write an editorial that includes citations to yours and other papers. Of course, the best position to be in is when you are one of the organising members. Your name will be associated with the special issue as well as your contribution. This means that when the special issue is published, each of the papers in it will already have citations leading to a positive [Matthew Effect](#matthew) from day one of publication.

Probably the easiest way of pulling together a special issue is to organise a symposium at a conference and ask those participating to contribute to the special issue. If this is done long enough in advance and you are good at organising then it can work very well. Talk to your [mentor](#mentor) about the possibility of their help in organising a symposium.

### Traditional Media 
To get your work cited you need to get people reading it. As scientists read the [traditional media](#pressrelease) it can be a way of bringing your work to their attention, and result in additional citations. There are other benefits of having your work highlighted by the traditional media. You should always make sure that you get a plug in for your institution and any prominent funders. There are now additional metrics (e.g. [Altmetrics](#altmetrics)) that track traditional and social media and in the future you might be able to use these altmetrics to your advantage in getting a job, a promotion or tenure.

### Not so legitimate ways of increasing citations {#notlegitimate}
As well as the legitimate ways of increasing citations and visibility of your work that I've listed above there are other illegitimate ways that I have heard about. I have heard of certain laboratories that have _quid pro quo_ arrangements where they agree to cite each other's publications [see @ritchie2020science].

Another way of increasing citations might be to have a co-author that is very well known. Certainly having a co-author that is very well known is likely to increase your citations but only if they actually do something to contribute to the work. See the section on [honorary authorship](#honorary).

@@

Journals have also been known to manipulate citations in order to increase their [Impact Factors](#impactfactor) and therefore their perceived level of quality. A group of physics journals from Romania were found to have clearly manipulated self-citations to increase their own impact factors [see @heneberg2016excessive]. There can also be a level of coercive pressure from journal editors to cite papers from their journals published within the two year citation window [@chorus2016large-scale]. Such practices are now systematically analysed by the larger literature databases, and clear levels of [Impact Factor](#impactfactor) manipulation result in delisting of the journal from the database.

@@

As you read through papers in your specialist area you will notice that there are some papers that seem to get cited over and over again as standard examples of a particular phenomenon. These are often some of the first examples published and have also been published in higher ranking journals. If you are the 20th person to have shown a particular phenomenon then it is unlikely that your paper is going to end up being cited the most.

There have been investigations into what makes certain papers more likely to be cited further. One of the factors that appears to be significant is the "small-study effect" - low precision studies that produce large effect sizes, either spuriously [i.e. a Type I error: @forstmeier2017detecting, or genuinely @fanelli2017meta-assessment].
